Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis

### 2023-03-31 analysis/paper/paper.Rmd
edit the intro

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000]. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4,700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015]. The Neolithic Hengduan Mountain regions are therefore essential for understanding the processes of agricultural expansion. Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as those grains found in Sichuan [@Huan2022; @Hao2022]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

To explore whether the new crops or species became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a]. The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016], and millet could have already become part of the local diet [@Lin2022]. Although some of these analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period. To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use? Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic period? We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.

The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies, or culinary practices across a region over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020]. Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers for tracing their original foodstuffs [@Copley2005; @Whelton2021]. This is due to the protection of clay matrices from microbiological degradation and water leaching, which allows lipid molecules to accumulate and be well preserved in the porous fabric of pottery over a thousand years [@Evershed1993; @Evershed2008; @Roffet2017]. In this study, we detected and identified lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020]. We attempt to investigate culinary practice during a period of frequent cultural interactions with agricultural expansion through our case study.

@@

```{r SW-china-map, fig.cap = "The archaeological sites, Guijiabao and Gaoshan, and other locations in southwest China that are mentioned in this study. Map data is from naturalearthdata.com"}
library(here)
knitr::include_graphics(here("analysis", "figures", "Sichuan-sites-map.png"))
```

<!--Yanyuan area and Guijiabao site-->
We focus on Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)). Guijiabao is unique for understanding the early adoption of agriculture due to its rarity, which contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone. For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles [@Hao2022].

>>>>

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas are usually accompanied by the exchange of goods, technology, or ideas that could have cultural or economic impacts on any people groups that are involved in the interactions that result from these movements [@Burmeister2000]. The Hengduan Mountain ranges and valleys in southwestern China are geographical areas that connect the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau and by which movement to these plateaus must occur. They play an important role in the cultural exchanges and spread of crops into southwestern China and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate on the routes of rice/millet dispersal from northwest China and the middle Yangtze valley to southwestern China [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015]. The Hengduan Mountain regions in the Neolithic period are therefore essential for understanding the processes of agricultural expansion. Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as the grains found in Sichuan [@Huan2022; @Hao2022]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

To explore whether new crops or terrestrial animals became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues within ceramic vessels from Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan that have been excavated so far [@Guijiabao2017a; @Guijiabao2020a]. The new evidence at Guijiabao reveals that millet and rice were both present there in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016], and millet could have already been part of the local diet [@Lin2022]. Although some of these analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period. To understand the extent of consuming new crops or terrestrial animals at Guijiabao, we ask the questions: What were the culinary practices that can be deduced by the pottery? Can we identify the new crops or terrestrial animals that were incorporated into the local food diets during the Neolithic period? We answer these questions by examining potsherds using organic residue analysis, which allows us to directly explore food cooked or processed in ceramic vessels.

The recent growing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies, or culinary practices across a region over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020]. Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers for tracing food origins [@Copley2005; @Whelton2021]. This is possible due to the protection of clay matrices from microbiological degradation and water leaching, which allows lipid molecules to accumulate and be well preserved in the porous fabric of pottery for over a thousand years [@Evershed1993; @Evershed2008; @Roffet2017]. In this study, we detected and identified lipid residues absorbed into the ceramic matrix that can reveal the original contents cooked or stored in a vessel during its lifetime of use [@Dudd1999; @Miller2020]. Through our case study, We attempt to investigate culinary practices during a period of frequent cultural interactions that accompanied agricultural expansion .

@@

```{r SW-china-map, fig.cap = "The archaeological sites Guijiabao and Gaoshan and other locations in southwest China that are mentioned in this study. Map data is from naturalearthdata.com"}
library(here)
knitr::include_graphics(here("analysis", "figures", "Sichuan-sites-map.png"))
```

<!--Yanyuan area and Guijiabao site-->
We focus on the site at Guijiabao located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)). The Guijiabao site allows us to understand the early adoption of agriculture because of its uniqueness and rarity: it contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022]. The chronology of the Guijiabao site can be divided into three temporal phases, the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone. For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles [@Hao2022].
